---
title: "Intro to Tidyverse: Data Wrangling"
output: html_notebook
---

#### Note

Column names must be a single word (or_use_underscores). They can't have spaces.

## Setup

```{r}
library(dplyr)
library(readr)
```


## Import data

```{r}
gapminder <- read_tsv("gapminder.tsv", col_names = TRUE)
```


## What type of data do we have?

```{r}
# gapminder is a data frame: data.frame
# specifically, a special data frame called a tibble
class(gapminder)
```


## Look at the structure

```{r}
str(gapminder)
```


## `filter()`

Filtering data gives you a subset of rows. 

```{r}
# fiter for year 2007
gapminder %>%
    filter(year == 2007)
```

Notice that we went from 1704 rows to 142 rows. Now let's filter by country.

```{r}
# filter by country
gapminder %>%
    filter(country == "United States")
```

We can also specify multiple conditions in a filter. Here, we only want the single observation for the year 2007 where the country is the United States.

```{r}
# get datapoint: year 2007, USA
gapminder %>%
    filter(year == 2007, country == "United States")
```


## `arrange()`

Arrange sorts the rows (observations) in a data set in ascending or descending order based on one of its columns (variables). You specify what column you want to arrange by.

```{r}
# sor in ascending order
gapminder %>%
    arrange(gdpPercap)
```

Say we wan to sort in descending order instead.

```{r}
# sort descending order
gapminder %>%
    arrange(desc(gdpPercap))
```

Say you wanted to find the highest GDP per capita countries in a single year. Combine `filter()` and `arrange()` to get your data.

```{r}
# choose observations only in 2007, rank highest GPD to lowest
# highest GDP per capita countries in 2007 
gapminder %>%
    filter(year == 2007) %>%
    arrange(desc(gdpPercap))
```

## `mutate()`

If you want to change one of the variables in your dataset based on the other ones or you want to add a new variable (column).

Information to the right of the equal sign is being evaluated. The variable on the let is what is being replaced.

```{r}
# reducing population to millions
gapminder %>%
    mutate(pop = pop / 1000000)
```

Now say we want to add a new variable. We have GDP per capita, which is GDP/population, but say we want total GDP.

```{r}
# create new GDP variable
gapminder %>%
    mutate(gdp = gdpPercap * pop)
```

## Combine the verbs: filter, arrange, mutate

```{r}
# countries with highest GDP in 2007
gapminder %>%
    mutate(gdp = gdpPercap * pop) %>%
    filter(year == 2007) %>%
    arrange(desc(gdp))
```

