---
title: "ggplot2 Visualizing Data"
output: html_notebook
---

## Setup

```{r}
library(dplyr)
library(ggplot2)
library(readr)
```

## Import data

```{r}
gapminder <- read_tsv("gapminder.tsv", col_names = TRUE)
```

When you're working on visualizing only a subset of data its useful to save that data as a new data frame.

```{r}
gapminder_07 <- gapminder %>%
    filter(year == 2007)
```

Suppose you want to examine the relationship between a country's wealth and its life expectancy. You could put GDP per capita on the x-axis and life expectancy on the y-axis


## Three parts to a ggplot graph

1. the data we're visualizing (gapminder_07)
2. mapping variables in your data to aesthetics in your graph
3. sepcify the type of graph you're creating by adding a geom layer (for example). Geom means you're adding a type of geometeric object to the graph. The point indicates a scatterplot.


```{r}
# relationship bewteen a country's wealth and life expectancy
ggplot(gapminder_07, aes(x = gdpPercap, y = lifeExp)) +
    geom_point()
```

## 1952 Examples

Filter gapminder for observations from the year 1952, and assign it to a new dataset gapminder_1952.

```{r}
gapminder_1952 <- gapminder %>%
    filter(year == 1952)
```

Change the scatter plot of gapminder_1952 so that (pop) is on the x-axis and GDP per capita (gdpPercap) is on the y-axis.

#### comparing population & GDP per capita

```{r}
# put pop on the x-axis and gdpPercap on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point()
```

#### comparing population & life expectancy

```{r}
# Create a scatter plot with pop on the x-axis and lifeExp on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) + 
    geom_point()
```

