---
title: "ggplot2 Visualizing Data"
output: html_notebook
---

## Setup

```{r}
library(dplyr)
library(ggplot2)
library(readr)
```

## Import data

```{r message=FALSE}
gapminder <- read_tsv("gapminder.tsv", col_names = TRUE)
```

When you're working on visualizing only a subset of data its useful to save that data as a new data frame.

```{r}
gapminder_07 <- gapminder %>%
    filter(year == 2007)
```

Suppose you want to examine the relationship between a country's wealth and its life expectancy. You could put GDP per capita on the x-axis and life expectancy on the y-axis


## Three parts to a ggplot graph

1. the data we're visualizing (gapminder_07)
2. mapping variables in your data to aesthetics in your graph
3. sepcify the type of graph you're creating by adding a geom layer (for example). Geom means you're adding a type of geometeric object to the graph. The point indicates a scatterplot.


```{r}
# relationship bewteen a country's wealth and life expectancy
ggplot(gapminder_07, aes(x = gdpPercap, y = lifeExp)) +
    geom_point()
```

The probelm with the above plot and some others below is the enormous distribution in scale of the x-axis, with some countries having GDP per capita in the tens of thousands while others are only in the hundreds. When you have a wide distribution like that many observations get crammed into the same area of the plot and can be hard to differentiate. In these graphs, we see it on the left side.


## 1952 Examples

Filter gapminder for observations from the year 1952, and assign it to a new dataset gapminder_1952.

```{r}
gapminder_1952 <- gapminder %>%
    filter(year == 1952)
```

Change the scatter plot of gapminder_1952 so that (pop) is on the x-axis and GDP per capita (gdpPercap) is on the y-axis.

#### comparing population & GDP per capita

```{r}
# put pop on the x-axis and gdpPercap on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point()
```

#### comparing population & life expectancy

```{r}
# Create a scatter plot with pop on the x-axis and lifeExp on the y-axis
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) + 
    geom_point()
```

## Log Scales

When you have data with a wide distribution range it's helpful to use a log scale on the axises. A log scale is a scale where each fixed distance represents a multiplication of the value. So a log-base-10 scale means that the second tick mark on the axis represents a 10x increase over the previous one. Think 10...100...1000...1000.

The plot below is the same data, but now each unit on the x-axis represents a change of 10 times the GDP. And we can see the relationship now looks more linear.

```{r}
ggplot(gapminder_07, aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  scale_x_log10()
```

## 1952 Examples: x-axis on Log10 scale

```{r}
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point() +
  scale_x_log10()
```

## 1952 Examples: both x and y axis on Log10 scale

```{r}
ggplot(gapminder_1952, aes(x = pop, y = gdpPercap)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

## Add color & size to plots

Continent is a categorical variable. A good way to represent categorical variables is using color. 

```{r}
# added color
ggplot(gapminder_07, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_x_log10()
```

Population is numerical value so a good way to represent it is in the size of the scattterplot points.

```{r}
# including population as size of points
ggplot(gapminder_07, aes(x = gdpPercap, y = lifeExp,
                         color = continent,
                         size = pop)) +
  geom_point() +
  scale_x_log10()
```

